---
title: "Discord Assignment - Week 8: Interpreting Slope Uncertainty"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Instructions

Please use the code chunks below to:  

1) load and clean
2) select two variables
3) bootstrap fit a linear model
4) calculate the 95% confidence interval of the slope coefficient

For your Discord post, please include a statement on:

	1.	your process (one sentence),
	2.	fitted slope + 95% bootstrap CI,
	3.	evaluate whether slope CI includes 0?
	4.	**if your CI includes 0** one reason the linear model might break (nonlinearity, confounding, diurnal structure, etc.).
	
	
I loaded and cleaned the dataset, selected "fc" as my response variable and "gpp" as my predictor variable, and then used bootstrap resampling to fit a linear model and produce a slope distribution (estimated). 

The fitted slope between "fc" and "gpp" was around -0.88 with a 95% bootstrap CI of [-0.892, -0.861]. The CI does not include a 0, meaning there is a strong negative relationship between the two variables. Specifically that as "gpp" increases, "fc" decreases in the sample. 
---

# Data

## Load the `US-AMS` dataset, clean names, filter for a few days in Spring 2023

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the CSV from the repo and clean column names.
# TODO: finish this pipeline.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023 & doy > 120 & doy < 180)

glimpse(df)
```

# Pick two variables

## Unlike labs, neither needs to be a flux variable (`fc`, `fch4`, `gpp`, `reco`)

```{r choose_vars, echo=TRUE, eval=FALSE}
response_var <- c(
  "fc"  # e.g., "fc"
)
predictor_var <- c(
    "gpp"  # e.g., "ta"
)
```

# Bootstrap fit a linear model between the two variables

```{r ex3_bootstrap_lm, echo=TRUE, eval=FALSE}
# fix to create smaller sample dataset
df <- df %>% slice_sample(n = 1000, replace = FALSE)

set.seed(480)

B <- 2000

boot_coef <- replicate(B, {

  samp_rows <- df %>% slice_sample(n = nrow(df), replace = TRUE)

  fit <- lm(
    samp_rows[[response_var]] ~ samp_rows[[predictor_var]]
  )

  coef(fit)

})

boot_coef <- t(boot_coef)
```

# Evaluate the CI slope range

```{r ex3_ci_params, echo=TRUE, eval=FALSE}
quantile(boot_coef[,2], probs = c(0.025, 0.975))  # slope CI
```
