---
title: "Lab 05 — Sampling and Bootstrap Distributions"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

In EAES we almost never observe an entire population. Instead, we take **samples** and compute **point estimates** (like a mean) to infer **population parameters**.

This lab focuses on four core ideas:

1. **Relative error** of point estimates (accuracy)
2. **Standard error (SE)** and how it changes with sample size
3. How **replication** (many repeated samples) creates an approximate **sampling distribution**
4. (Intro) **Bootstrap** distributions: resampling from one sample to approximate uncertainty

## Coding expectations (Week 7+)

This lab uses a mix of:
- **Fill-in-the-blank** (quick checks)
- **Partial pipelines** (you finish the analysis)
- **From-scratch chunks** (you write the full solution)

Use slides + past labs + Discord as needed, but make sure your final code is **your own** and runs **top-to-bottom**.

---

# Data

## Load data (partial pipeline)

Complete the pipeline so `df` is a clean tibble. Filter for 2023, and replace fill values (-9999) with na.

```{r load_data, echo=TRUE, eval=T}
# GOAL: Read the CSV from the repo and clean column names.
# TODO: finish this pipeline.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  filter(Year_local == 2023)

glimpse(df)
```

## Derive time columns 

Write code to add:
- `date` from `year_local` + `doy`
- `month` as labeled abbreviated month (Jan–Dec)
- `day_night` as "Day"/"Night" from `daytime` (0/1)

```{r derive_time, echo=TRUE, eval=T}
# GOAL: Create date/month/day_night columns.
# HINT: DOY is 1-indexed, so use (doy - 1) with origin "YYYY-01-01".

df <- df %>%
 mutate(
    # TODO: create a Date column from year_local and doy
    # HINT: Jan 1 is DOY = 1, so use (doy - 1) with origin = "YYYY-01-01"
    date = as.Date(DOY - 1, origin = paste0(Year_local, "-01-01")),

    # TODO: create a month column (numeric 1–12 or labeled months)
    month = month(date, label = TRUE, abbr = TRUE),

    # TODO: make a day/night label using daytime (0/1)
    day_night = if_else(daytime == 1, "Day", "Night")
  )

df %>% count(month)
df %>% count(day_night)
```

---

# Population parameters vs point estimates

## 1) Choose variables (fill-in)

Pick **2–3 variables** to analyze. Include at least **one flux** variable (e.g., `nee`, `fc`, `gpp`) and optionally one meteorological variable (e.g., `ta`, `le`).

```{r choose_vars, echo=TRUE, eval=T}
vars <- c(
  "FC",  # e.g., "fc"
  "TA",  # e.g., "ta"
  "GPP"   # optional third variable
)

# CHECK: confirm these columns exist (should return character(0))
setdiff(vars, names(df))
```

---

## 2) Compute population parameters (partial pipeline)

Complete the code to compute **population mean** and **population SD** for each selected variable.

Your output should be a tibble named `pop` with columns: `var`, `mean`, `sd`.

```{r pop_params, echo=TRUE, eval=T}
# GOAL: Create a tidy table of population parameters for vars.
# TODO: complete this pipeline.

pop <- df %>%
  summarise(across(
    all_of(vars),
    list(
      mean = ~mean(.x, na.rm = TRUE),
      sd   = ~sd(.x, na.rm = TRUE)
    ),
    .names = "{.col}__{.fn}"
  )) %>%   
  pivot_longer(
    everything(),
    names_to = c("var", "stat"),
    names_sep = "__",
    values_to = "value"
  ) %>%   
  pivot_wider(names_from = stat, values_from = value)       # TODO: pivot_wider

pop
```

**Prompt (2–3 sentences):** Which variable has the largest population SD? What does that imply about the variability of that process?

The GPP variable has the largest standard deviation. This means that process has the most variability. 

---

# Relative error for point estimates (accuracy)

We will estimate the population mean with a sample mean, then compute **relative error**:

\[
\text{Relative error (\%)} = 100 \times \frac{|\mu - \bar{x}|}{|\mu|}
\]

## 3) One sample: compute relative error 

Write code to:

1. set a seed
2. choose a sample size `n_samp`
3. draw a simple random sample `samp`
4. compute `sample_mean` for each variable in `vars`
5. join to population means from `pop`
6. compute `rel_error_pct` for each variable

Return a tibble with columns: `var`, `sample_mean`, `pop_mean`, `rel_error_pct`.

```{r one_sample_relerr, echo=TRUE, eval=T}
# GOAL: Relative error for point estimates from one random sample.

set.seed(123)
n_samp <- 100

samp <- df %>%
  slice_sample(n = n_samp)

# get mean point estimate for each variable
est <- samp %>%
  summarise(across(all_of(vars), ~mean(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "sample_mean")

# compute the relative error for each variable
relerr <- est %>%
  left_join(pop %>% select(var, pop_mean = mean), by = "var") %>%
  mutate(rel_error_pct = 100 * abs(pop_mean - sample_mean) / abs(pop_mean)) %>%
  arrange(desc(rel_error_pct))

relerr
```

**Prompt (3–4 sentences):** Which variable had the largest relative error at your chosen `n_samp`?  
Give a scientific reason why some variables are harder to estimate from a small sample.

FC had the largest relative error at the chosen `n_samp`. Flux variables are going to show more variability compared to variables like temperature. They fluctuate rapidly and may include extreme values, small samples are less representative of the full distribution. 

---

## 4) Relative error vs sample size (partial code + you complete)

We will focus on **one variable** and simulate many repeated studies. The next code chunk is doing quite a lot. Please review the #comments throughout and check you understand generally what each section is doing.

```{r relerr_by_n, echo=TRUE, eval=T}
set.seed(480)

n_grid <- c(25, 100, 300, 1000)   # e.g., c(25, 100, 300, 1000)
reps <- 500                        # e.g., 500 or 1000

target_var <- "FC"                # e.g., "fch4" or "fc"

# Pull the population vector (remove missing values once, up front)
x_pop <- df[[target_var]]
x_pop <- x_pop[!is.na(x_pop)]

# Population mean for the chosen variable
pop_mean_target <- pop %>%
  filter(var == target_var) %>%
  pull(mean)

# ---- Simulation using explicit for loops ----
sim_list <- vector("list", length(n_grid))

for (i in seq_along(n_grid)) {

  n_samp <- n_grid[i]

  # store relative error from each repeated "study"
  relerrs <- numeric(reps)

  for (j in seq_len(reps)) {

    x_samp <- sample(x_pop, size = n_samp, replace = FALSE)
    xbar <- mean(x_samp)

    relerrs[j] <- 100 * abs(pop_mean_target - xbar) / abs(pop_mean_target)
  }

  sim_list[[i]] <- tibble(
    n = n_samp,
    rel_error_pct = relerrs
  )
}

sim_relerr <- bind_rows(sim_list)

sim_relerr
```

### Summarize + plot (from scratch)

1) Summarize the simulated relative error by `n` using:
- median
- 90th percentile

2) Make a plot that communicates “relative error decreases with n”.

```{r summarize_plot_relerr, echo=TRUE, eval=T}
# GOAL: Summarize and plot the relationship between n and relative error.
# TODO: write your own code.

summary_relerr <- sim_relerr %>%
  group_by(n) %>%
  summarise(median_rel_error = median(rel_error_pct),
            p90_rel_error = quantile(rel_error_pct, 0.9))
  
summary_relerr

p_relerr <- sim_relerr %>%
  ggplot(aes(x = factor(n), y = rel_error_pct)) +
  geom_boxplot()+
  
  labs( x = "Sample size (n)", y = "Relative error (%)")

p_relerr
```

**Prompt (3–4 sentences):** Describe the pattern. Is improvement from n=25→100 larger than 300→1000?  
Why does accuracy show “diminishing returns” as n increases?

Relative error decreases as sample size increases. The improvement from n = 25 to 100 is much larger than from n = 300 to 100. This pattern reflects lower improvement because sampling variability shrinks at a specific rate. As n becomes large, each observation reduces error by smaller amounts. 

---

# Standard error and sampling distributions

For the sample mean, the standard error is approximately:

\[
SE(\bar{x}) \approx \frac{\sigma}{\sqrt{n}}
\]

where \(\sigma\) is population SD.

## 5) Theory vs simulation (partial pipeline)

Compute:
- theoretical SE
- simulated SD of sample means (over many replications)

```{r se_compare, echo=TRUE, eval=T}
set.seed(480)

# TODO: pick a variable and sample size
target_var <- "GPP"
n_samp <- 100
reps <- 1000

pop_sd_target <- pop %>% filter(var == target_var) %>% pull(sd)
pop_mean_target <- pop %>% filter(var == target_var) %>% pull(mean)

means <- replicate(
  reps,
  mean(sample(df[[target_var]], size = n_samp, replace = FALSE), na.rm = TRUE)
)

se_theory <- pop_sd_target / sqrt(n_samp)
se_sim    <- sd(means)

tibble(
  target_var = target_var,
  n = n_samp,
  se_theory = se_theory,
  se_simulated = se_sim
)
```

### Plot the sampling distribution (fill-in)

```{r plot_sampling_dist, echo=TRUE, eval=T}
tibble(mean_est = means) %>%
  ggplot(aes(x = mean_est)) +
  geom_histogram(bins = 20, alpha = 0.85) +
  geom_vline(xintercept = pop_mean_target, linetype = "dashed", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Sample mean of ", target_var),
    y = "Count",
    title = "Sampling distribution of the mean",
    subtitle = paste0("n = ", n_samp, "; reps = ", reps, " (dashed = population mean)")
  )
```

**Prompt (3–5 sentences):** How close were `se_theory` and `se_simulated`?  
What changes if you increase `n_samp` vs increase `reps`?

If you increase `n_samp` SE decreases because variability of the mean shrinks with larger samples. If you increase `reps` SE does not change. It does however make the estimate of the sampling distribution more precise. 

---

# Bootstrap intuition (one sample, many resamples)

Bootstrap resampling approximates uncertainty **when you only have one sample**.

Key idea: treat your sample as a stand-in for the population and resample **with replacement**.

## 6) Bootstrap distribution of the mean 

Write code to:

- take ONE sample of size `n_samp` from `df`
- generate `B` bootstrap resamples (with replacement) from that sample
- compute bootstrap means
- compute bootstrap SD
- plot the bootstrap distribution

```{r bootstrap_mean, echo=TRUE, eval=T}
set.seed(480)

# TODO: choose settings
target_var <- "TC"
n_samp <- 100
B <- 1000

# one "field campaign" sample
samp <- tibble(x = sample(x_pop, size = n_samp, replace = FALSE))

# bootstrap sampling of the mean
boot_means <- replicate(
  B,
  mean(sample(samp$x, size = n_samp, replace = TRUE))
)

boot_sd <- sd(boot_means)

tibble(
  sample_mean = mean(samp$x),
  boot_sd = boot_sd
)
```

```{r plot_boot, echo=TRUE, eval=T}
# TODO: plot boot_means and mark the sample mean
tibble(boot_mean = boot_means) %>%
ggplot(aes(boot_mean)) +
  geom_histogram(bins = 20, alpha = 0.85) +
  geom_vline(xintercept = mean(samp$x)) +
  labs(x = "Bootstrap means", y = "Count")
```

**Prompt (4–6 sentences):** Compare the bootstrap distribution to the sampling distribution from Part III.  
What is the bootstrap trying to approximate? When is bootstrap especially useful in EAES?

The bootstrap distribution is trying to approximate the sampling distribution of the mean. The sampling distribution was formed using repetitive sampling of the population, the bootstrap resamples from one sample. It is trying to approximate the variability we would see if we had repeatedly sampled from the population. Bootstrap is useful when full population data is unavailable but we need uncertainty estimates. 

---


# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` to your GitHub repo.
